Repeated loss of variation in insect ovary morphology highlights the role of developmental constraint in life-history evolution

### 2021-01-19 manuscript.Rmd
updating Rmd and manuscript files in association with submission

---
title: Repeated loss of variation in insect ovary morphology highlights the role of developmental constraint in life-history evolution
output:
  bookdown::pdf_document2:
  keep_tex: yes
  latex_engine: pdflatex
  #bookdown::html_document2: default
toc: false
bibliography: manuscript_supplement.bib
csl: nature.csl
link-citations: yes
header-includes:
 \usepackage{xr} \externaldocument{manuscript_supplement}
 \usepackage{float}
 \usepackage{array}
 \usepackage{xcolor,colortbl}
 \floatplacement{figure}{H}
---

@@

To evaluate the robustness of our results to uncertainty in the phylogenetic relationships, all Phylogenetic Generalized Least Squares (PGLS) analyses were performed 1000 times over a posterior distribution of trees, using a Brownian Motion based covariance matrix in the R package ape (version `r packageVersion('ape')`)[@paradis2004ape] and nlme (version `r packageVersion('nlme')`)[@pinheiro2014r]. For regressions at the species and genus level, we reshuffled and matched records for each iteration to account for variation across records for the same taxon. For regressions at the family level we recalculated the average ovariole number per insect family, randomly downsampling the representation for each family by half. To weight traits by body size, we calculated the phylogenetic residuals[@revell2010phylogenetic] of each trait to body size, and then compared the evolution of these residuals using a PGLS regression. See supplementary methods section \ref{sec:PGLS} for details.

@@

Other comparisons of model fit were performed using the R package geiger(version `r packageVersion('geiger')`)[@harmon2007geiger] and validated using a parametric bootstrap with the R package arbutus (version `r packageVersion('arbutus')`)[@pennell2015model]. See supplementary methods section \ref{sec:bootstrap}.

Analyses of evolutionary rate were performed using BAMM (version 2.5.0)[@rabosky2014automatic]. For this analysis, we calculated the average ovariole number (log~10~ transformed) for each genus present in the phylogeny (507 taxa). We used the R package BAMMtools (version `r packageVersion('BAMMtools')`)[@rabosky2014bamm] to select appropriate priors, and ran BAMM for the maximum number of generations ($2*10^{-9}$), sampling every 10^6^ generations. Convergence was evaluated both visually (Fig. \ref{fig:BAMM-analysis-convergence}) and numerically. Running BAMM for the maximum possible number of generations and selecting the optimum burn-in (Fig. \ref{fig:BAMM-analysis-burnin}) resulted in an effective size for the number of shifts of `r round(es_nshifts,2)`, and for log-likelihood of `r round(es_logl,2)`. Repeated BAMM analyses showed similar distributions of high and low rate regimes, indicating the implications for ovariole number evolution are robust to uncertainty in rate estimates. See supplementary methods section \ref{sec:BAMM} for details.

@@

The code and phylogenetic trees required to reproduce all the analyses, figures, and generate the manuscript files are provided at 'https://github.com/shchurch/insect_ovariole_number_evolution_2020', commit 6cf446a. All analyses performed in R (version 4.0.3) were done so in a clean environment, built with conda (version 4.9.2), and instructions for rebuilding this environment are provided in the same repository.

@@

We further explored these results using two methods: First, to evaluate our findings against alternative evolutionary hypotheses, we compared these results to regressions based on simulated data. Our results showed that when considering body size, the slope of the regression of egg size and ovariole number is lower than we would expect to observe by chance, as assessed by comparing to data simulated with no evolutionary correlation (Fig. \ref{fig:simulated}). However, for both adult body length and dry mass, the slope of the regressions on observed data are not within the range that would be expected under an strong negative correlation (slope of -1 in log-log space, Fig. \ref{fig:simulated}). This suggests the presence of a weak evolutionary relationship bewteeen ovariole number and egg size, when accounting for body size.

@@

Using the dataset compiled here and a previously published phylogeny of insects (Fig. 5a)[@church2019insect], we modeled the rate of evolutionary change in ovariole number (Figs. \ref{fig:arbutus}, \ref{fig:BAMM-analysis-convergence}, \ref{fig:BAMM-analysis-burnin}, \ref{fig:BAMM-analysis-results}). We observed substantial rate heterogeneity in the evolution of ovariole number (Fig. \ref{fig:BAMM-analysis-results}), meaning that for some lineages ovariole number has evolved rapidly where in others, ovariole number has evolved very slowly or not at all. The most striking example of this are the multiple lineages which have independently evolved invariant or near-invariant ovariole number across taxa (e.g. nearly all Lepidoptera have exactly eight ovarioles, Fig. 5b, cyan), from an ancestral variable state. These invariant lineages were identified by finding regions of the phylogeny that experience extremely low rates of ovariole number diversification (Figs. \ref{fig:BAMM-analysis-results}, \ref{fig:BAMM-anlaysis-mean-rates}). Using this approach, we found that invariant ovariole numbers have evolved at least nine times independently across insects, with several subsequent reversals from invariant to variable states (Fig. 5a).

@@

SHC led the data collection, analysis, and writing of the manuscript. NM performed the initial literature search, data collection, and analysis. BASdM assembled all phylogenies for analysis. SD, BASdM, and CGE contributed to data analysis, visualization, discussion, and writing.

>>>>

---
title: Repeated loss of variation in insect ovary morphology highlights the role of developmental constraint in life-history evolution
output:
  #word_document
  bookdown::pdf_document2:
  keep_tex: yes
  latex_engine: pdflatex
  #bookdown::html_document2: default
toc: false
bibliography: manuscript_supplement.bib
csl: nature.csl
link-citations: yes
header-includes:
 \usepackage{xr} \externaldocument{manuscript_supplement}
 \usepackage{float}
 \usepackage{array}
 \usepackage{xcolor,colortbl}
 \floatplacement{figure}{H}
---

@@

To evaluate the robustness of our results to uncertainty in the phylogenetic relationships, all Phylogenetic Generalized Least Squares (PGLS) analyses were performed 1000 times over a posterior distribution of trees, using a Brownian Motion based covariance matrix in the R package ape (version `r packageVersion('ape')`)[@paradis2004ape] and nlme (version `r packageVersion('nlme')`)[@pinheiro2014r]. For regressions at the species and genus level, we reshuffled and matched records for each iteration to account for variation across records for the same taxon. For regressions at the family level, we recalculated the average ovariole number per insect family, randomly downsampling the representation for each family by half. To weight traits by body size, we calculated the phylogenetic residuals[@revell2010phylogenetic] of each trait to body size, and then compared the evolution of these residuals using a PGLS regression. See supplementary methods section \ref{sec:PGLS} for details.

@@

Other comparisons of model fit were performed using the R package geiger (version `r packageVersion('geiger')`)[@harmon2007geiger] and validated using a parametric bootstrap with the R package arbutus (version `r packageVersion('arbutus')`)[@pennell2015model]. See supplementary methods section \ref{sec:bootstrap}.

Analyses of evolutionary rate were performed using BAMM (version 2.5.0)[@rabosky2014automatic]. For this analysis, we calculated the average ovariole number (log~10~ transformed) for each genus present in the phylogeny (507 taxa). We used the R package BAMMtools (version `r packageVersion('BAMMtools')`)[@rabosky2014bamm] to select appropriate priors, and ran BAMM for the maximum number of generations ($2*10^{9}$), sampling every 10^6^ generations. Convergence was evaluated both visually (Fig. \ref{fig:BAMM-analysis-convergence}) and numerically. Running BAMM for the maximum possible number of generations and selecting the optimum burn-in (Fig. \ref{fig:BAMM-analysis-burnin}) resulted in an effective size for the number of shifts of `r round(es_nshifts,2)`, and for log-likelihood of `r round(es_logl,2)`. Repeated BAMM analyses showed similar distributions of high and low rate regimes, indicating the implications for ovariole number evolution are robust to uncertainty in rate estimates. See supplementary methods section \ref{sec:BAMM} for details.

@@

The code and phylogenetic trees required to reproduce all the analyses, figures, and generate the manuscript files are provided at 'https://github.com/shchurch/insect_ovariole_number_evolution_2020', commit 9876387. All analyses performed in R (version 4.0.3) were done so in a clean environment, built with conda (version 4.9.2), and instructions for rebuilding this environment are provided in the same repository.

@@

We further explored these results using two methods: First, to evaluate our findings against alternative evolutionary hypotheses, we compared these results to regressions based on simulated data. Our results showed that when considering body size, the slope of the regression of egg size and ovariole number is lower than we would expect to observe by chance, as assessed by comparing to data simulated with no evolutionary correlation (Fig. \ref{fig:simulated}). However, for both adult body length and dry mass, the slope of the regressions on observed data are not within the range that would be expected under a strong negative correlation (slope of -1 in log-log space, Fig. \ref{fig:simulated}). This suggests the presence of a weak evolutionary relationship between ovariole number and egg size, when accounting for body size.

@@

Using the dataset compiled here and a previously published phylogeny of insects (Fig. 5a)[@church2019insect], we modeled the rate of evolutionary change in ovariole number (Figs. \ref{fig:arbutus}, \ref{fig:BAMM-analysis-convergence}, \ref{fig:BAMM-analysis-burnin}, \ref{fig:BAMM-analysis-results}). We observed substantial rate heterogeneity in the evolution of ovariole number (Fig. \ref{fig:BAMM-analysis-results}), meaning that for some lineages ovariole number has evolved rapidly where in others, ovariole number has evolved very slowly or not at all. The most striking example of this are the multiple lineages which have independently evolved invariant or near-invariant ovariole number across taxa (e.g., nearly all Lepidoptera have exactly eight ovarioles, Fig. 5b, cyan), from an ancestral variable state. These invariant lineages were identified by finding regions of the phylogeny that experience extremely low rates of ovariole number diversification (Figs. \ref{fig:BAMM-analysis-results}, \ref{fig:BAMM-anlaysis-mean-rates}). Using this approach, we found that invariant ovariole numbers have evolved at least nine times independently across insects, with several subsequent reversals from invariant to variable states (Fig. 5a).

@@

SHC led the data collection, analysis, and writing of the manuscript. NM performed the initial literature search, data collection, and analysis. BASM assembled all phylogenies for analysis. SD, BASM, and CGE contributed to data analysis, visualization, discussion, and writing.
